---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# latentClass

<!-- badges: start -->
<!-- badges: end -->

latentClass is an R package that provides a vanilla implementation of latent class models with continuous variables (Gaussian) and categorical variables.

latentClass tries to identify subgroups of individuals within the data set. The
central assumption is that, once we know the class an individual is part of,
all items are independent (local independence). The likelihood of the model
is given by:

$$L = \prod_{i=1}^N \prod_{c=1}^C \pi_cp(x_{i1}|c)p(x_{i2}|c)\cdots p(x_{iP}|c),$$

where $N$ is the sample size, $C$ is the number of classes, $\pi_c$ is the
probability of class c, $p(x_{i1}|c)$ is the likelihood of observing
$x_{i1}$ given that person $i$ is in class $c$, and $P$ is the number of items.

Currently, latentClass supports normal (Gaussian) items and categorical items:

- Gaussian items: The likelihood for normal (Gaussian) items is given by $p(x) = \frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ (see https://en.wikipedia.org/wiki/Normal_distribution for more details).
- Categorical items: The likelihood for categorical items is given by $p(x) = p_1^{[x=1]}p_2^{[x=2]}\cdots p_k^{[x=k]}$ (see https://en.m.wikipedia.org/wiki/Categorical_distribution for more details).

Parameters are estimated with an Expectation Maximization optimizer. Additionally,
latentClass supports sample weights (see Murphy & Scrucca, 2012).

latentClass is one of multiple packages that implement latent class models
in R. The implementation of categorical items is heavily inspired by the poLCA
package (Drew et al., 2011). Similarly, the implementation of Gaussian items
is inspired by mclust (Scrucca, et al., 2023).
The implementation of the Expectation Maximization
optimizer follows that found in Blume (2002). Overall latentClass is a very
vanilla implementation of latent class models, lacking many of the more advanced
features of other packages (e.g., no latent class regressions).

## Installation

You can install the development version of latentClass from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("jhorzek/latentClass")
```

## Example

```{r}
library(latentClass)
data <- simulate_latent_class_data()
model <- latentClass(data = data,
                    # Define the number of classes:
                    n_classes = 3,
                    # Specify which of the items are categorical:
                    categorical = categorical(items = c("cat_1", "cat_2")),
                    # Specify which of the items are Gaussian:
                    normal = normal(items = c("norm_1", "norm_2")))

summary(model)
```

## References:

- Blume, M. (2002). Expectation maximization: A gentle introduction. Technical University of Munich Institute for Computer Science.
- Drew A. Linzer, Jeffrey B. Lewis (2011). poLCA: An R Package for Polytomous Variable
Latent Class Analysis. Journal of Statistical Software, 42(10), 1-29. URL
https://www.jstatsoft.org/v42/i10/
- Murphy, T. B., & Scrucca, L. (2012). Using Weights in mclust.
- Scrucca L, Fraley C, Murphy TB, Raftery AE (2023). _Model-Based Clustering,
Classification, and Density Estimation Using mclust in R_. Chapman and Hall/CRC. ISBN
978-1032234953, doi:10.1201/9781003277965 <https://doi.org/10.1201/9781003277965>,
<https://mclust-org.github.io/book/>